On Similarity

### 2016-04-28 paper.tex
another revision

In recent years, similarity received a lot of attention as an intrinsic evaluation of meaning representations \cite{mikolov2013efficient,Bruni:2012:DST:2390524.2390544}, because in a complete text processing system it is difficult to measure performance of a single (similarity) component. As a result, a new generation of datasets appeared from the computational linguistics community, for example, MEN \cite{Bruni:2012:DST:2390524.2390544} and SimLex-999 \cite{hill2014simlex}, that surpasses older datasets stemmed in psychology (refer to \newcite{1986-13502-00119860101} for a comprehensive list of datasets) and information retrieval (most notably \newcite{2002:PSC:503104.503110}) in quantity and attention to the evaluated relation.

Despite improvement, current similarity datasets deviate from the psychological understanding of similarity. Intuitively, similarity is the amount of common properties between entities, and is implemented as a contrast model where ``the similarity between objects is expressed as a function of their common and distinctive features'' \cite{Tversky1977}.

Categories are determined in terms of focal points, for example fruit names are arranged around the word \textit{fruit} in such a way that it is their nearest neighbor \cite{1986-13502-00119860101}. It is extremely important for a good semantic model to capture this property. Unfortunately, word pair similarity judgements do not capture this, as it is possible to arrange all fruit names close to \textit{fruit} but \textit{fruit} would not be the most common nearest neighbor. Moreover, as \newcite{turney2012domain} points out, categories that emerge from similarity judgements are different from lexical ontologies such as WordNet \cite{Miller:1995:WLD:219717.219748}, for example, \textit{traffic} and \textit{water} might be considered to be similar because of functional similarity exploited in hydrodynamic models of traffic, but their lowest common ancestor in WordNet is \textit{entity}.

>>>>

In recent years, similarity received a lot of attention as an intrinsic evaluation of meaning representations \cite{mikolov2013efficient,Bruni:2012:DST:2390524.2390544}, because in a complete text processing system it is difficult to measure performance of a single (similarity) component. As a result, a new generation of similarity datasets appeared from the computational linguistics community. The two notable datasets are MEN \cite{Bruni:2012:DST:2390524.2390544} and SimLex-999 \cite{hill2014simlex}. They surpasses older datasets stemmed in psychology (refer to \newcite{1986-13502-00119860101} for a comprehensive list of datasets) and information retrieval (for example \newcite{2002:PSC:503104.503110}) in quantity and attention to the evaluated relation. Both of them consist of word pairs assigned similarity (relatedness) scores based on human judgements.

Despite improvement, these datasets deviate from the psychological understanding of similarity. Intuitively, similarity is the amount of common properties between entities, and is implemented as a contrast model where ``the similarity between objects is expressed as a function of their common and distinctive features'' \cite{Tversky1977}.

One of the uncaptured properties of human similarity judgements is that categories are formed around \emph{focal points} \cite{1986-13502-00119860101}. Consider fruit names assessed for similarity. They will be arranged around the word \textit{fruit} in such a way that it is their nearest neighbor, making \textit{fruit} the focal point of the category. It is extremely important for a good semantic model to capture this property. Unfortunately, word pair similarity judgements do not capture this, as it is possible to arrange all fruit names close to \textit{fruit} but \textit{fruit} would not be the most common nearest neighbor. Moreover, as \newcite{turney2012domain} points out, categories that emerge from similarity judgements are different from lexical ontologies such as WordNet \cite{Miller:1995:WLD:219717.219748}, for example, \textit{traffic} and \textit{water} might be considered to be similar because of functional similarity exploited in hydrodynamic models of traffic, but their lowest common ancestor in WordNet is \textit{entity}.
